---
title: GSE96058 R Data Package
output: github_document   ### will keep md file!!!
bibliography: docs/_bibFiles/bibFile.bib
csl: docs/_bibFiles/cell-numeric.csl
link-citations: true
---

---

  [![Travis build status](https://travis-ci.com/12379Monty/GSE96058.svg?branch=master)](https://travis-ci.com/12379Monty/GSE96058)


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


This package contains data from NCBI GEO 
[Series GSE96058](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96058).
See  Breffer at al. (2018) [@Brueffer:2018aa].
### UP TO HERE
Data methylation profiles were collected to investigate
genome-wide mapping of 5-hydroxymethylcytosines
in circulating cell-free DNA as reported in Cai et al. (2019)
(https://gut.bmj.com/content/68/12/2195).

<!-- [@Cai:2019aa] pandoc citations dont work on github (github_document?) -->



## Data sets  

```{r dataSets, eval=F}
  library(GSE96058)

  tmp <- data(package='GSE96058')
  knitr::kable(tmp$results[, c("Item", "Title")])

```



## Installation


``` r
# install.packages("devtools")
devtools::install_github("12379Monty/GSE96058")
```

## Example


```{r example, eval=F}
library(magrittr)
library(GSE96058)
```

<br/>

* sampType is used to distinguish blood from tissue samples.  The latter
can be tumor or tumor adjacent (sampType==TI).  

```{r, eval=F}
with(sampDesc, table(sampType, trainValGroup, exclude=NULL))
```

<br/>

*  For blood samples, outcome is specific Dx which can be used to
investigate 5hmc profile differences between healthy, benign
and malignant samples.

```{r, eval=F}
with(sampDesc %>% dplyr::filter(sampType=='blood'), 
     table(outcome, trainValGroup, exclude=NULL))
```

<br/>

* outcome2 pools Benign+Healthy and Cirrhosis+CHB.  

```{r, eval=F}
with(sampDesc %>% dplyr::filter(sampType=='blood'), 
     table(outcome2, trainValGroup, exclude=NULL))
```

<br/>

* outcome3 is HCC vs the rest.  

 ```{r, eval=F}
with(sampDesc %>% dplyr::filter(sampType=='blood'), 
     table(outcome3, trainValGroup, exclude=NULL))
```

<br/>

* sampDesc is used to provide sample information 
for the feature count data.  

```{r, eval=F}
TUTI_samdesc <- sampDesc[colnames(TUTI_featureCount),]

with(TUTI_samdesc, table(sampType, tissueSubtype,exclude=NULL))

par(mfrow=c(2,1), mar=c(4,3,2,3),oma=c(0,0,2,0))
for(ST in unique(TUTI_samdesc$sampType)){
  ST_ndx <- which(TUTI_samdesc$sampType==ST)
  boxplot(log2(TUTI_featureCount+1)[,ST_ndx], outline=F, las=2, 
          ylab='log2(count+1)', xaxt='n')
  axis(side=1, outer=F, at=1:length(ST_ndx), 
       label=TUTI_samdesc$title[ST_ndx],las=2)
  title(paste('sampType =', ST)) 

  sample_median <- apply(log2(TUTI_featureCount[,ST_ndx]+1),2,median)
  abline(h=median(sample_median), col='green')
  axis(side=4, at=round(median(sample_median),1),las=2, 
       col.axis='green', line=-0.5,tick=F)
 
}
 

```

## Package page

* https://12379monty.github.io/GSE96058/

<!-- DOESNT WORK WITH GITHUB
## References
<div id="refs"></div>
-->

-----

Please note that the 'GSE96058' project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

<!--

Rscript -e "rmarkdown::render('README.Rmd')"

# nohup Rscript -e "rmarkdown::render('README.Rmd')" > README.log  &


-->
